Mixing Time Estimation in Reversible Markov Chains from a Single Sample Path

### 2015-10-31 arxiv-v2/empirical.tex
.
corner cases

  \STATE Empirical bounds for $|\wh{P}_{i,j}-P_{i,j}|$ for each $(i,j)
  \in [d]^2$:
  let $\emptail
    := \inf
    \braces{
      t\geq0 :
      2d^2 \parens{ 1 + \ceil{\ln(2n/t)} } e^{-t} \leq \delta
    }$ and
  \begin{equation}
    \wh{B}_{i,j}
    :=
    \sqrt{\frac{9e\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{2N_i}}
    + \frac{(3e+2.5)\emptail+1.5\abs{(1/d)-\wh{P}_{i,j}}}{N_i}
    .
    \notag
  \end{equation}
  \label{step:P-bound}

@@

It is easy to combine \cref{thm:err,thm:empirical} to yield intervals
whose widths shrink at at least as fast as both the non-empirical
intervals from \cref{thm:err} and the empirical intervals from
\cref{thm:empirical}.
Specifically, determine lower bounds on $\pimin$ and $\gap$ using
\cref{alg:empest},
\[
  \pimin \geq \min_{i \in [d]} \brackets{ \hat\pi_i - \hat{b} }_+
  , \quad
  \gap \geq \brackets{ \hatgap - \hat{w} }_+
  ;
\]
then plug-in these lower bounds for $\pimin$ and $\gap$ in the
deviation bounds in \cref{eq:gapbound} from \cref{thm:err}.
This yields a new interval centered around the estimate of $\gap$ from
\cref{thm:err}, and it no longer depends on unknown quantities.
The interval is a valid $1-2\delta$ probability confidence interval
for $\gap$, and for sufficiently large $n$, the width shrinks at the
rate given in \cref{eq:gapbound}.
We can similarly construct an empirical confidence interval for
$\pimin$ using \cref{eq:piminbound}, which is valid on the same
$1-2\delta$ probability event.\footnote{%
  For the $\pimin$ interval, we only plug-in lower bounds on $\pimin$
  and $\gap$ only where these quantities appear as $1/\pimin$ and
  $1/\gap$ in \cref{eq:piminbound}.
  It is then possible to ``solve'' for observable bounds on $\pimin$.
  See \cref{app:combined} for details.%
}
Finally, we can take the intersection of these new intervals with the
corresponding intervals from \cref{alg:empest}.
This is summarized in the following \lcnamecref{thm:combined}, which
we prove in \cref{app:combined}.
\begin{theorem}
  \label{thm:combined}
  The following holds under the same conditions as
  \cref{thm:empirical}.
  For any $\delta \in (0,1)$, the confidence intervals $\wh{U}$ and
  $\wh{V}$ described above for $\pimin$ and $\gap$, respectively,
  satisfy $\pimin \in \wh{U}$ and $\gap \in \wh{V}$ with probability
  at least $1-2\delta$.
  Furthermore, the widths of these intervals almost surely satisfy
  (as $n \to \infty$) 
  \[
    |\wh{U}|
    =
    O\Parens{
      \sqrt{\frac{\pimin\log\frac{d}{\pimin\delta}}{\gap n}}
    }
    ,
    \quad
    |\wh{V}|
    =
    O\Parens{
      \min\Braces{
        \sqrt{\frac{\log\frac{d}{\delta}\cdot\log(n)}{\pimin\gap n}}
        ,\,
        \hat{w}
      }
    }
  \]
  where $\hat{w}$ is the width from \cref{alg:empest}.
\end{theorem}

>>>>

  \STATE Empirical bounds for $|\wh{P}_{i,j}{-}P_{i,j}|$ for $(i,j){\in}[d]^2$:
  $c:=1.01$,
  $\emptail
    := \inf
    \braces{
      t\geq0 :
      2d^2 \parens{ 1 + \ceil{\log_c\frac{2n}{t}}_+ } e^{-t} \leq
      \delta
    }$,
  \begin{equation}
    \text{and} \quad
    \wh{B}_{i,j}
    :=
    \Parens{
      \sqrt{\frac{c\emptail}{2N_i}}
      + \sqrt{
        \frac{c\emptail}{2N_i}
        + 
        \sqrt{\frac{2c\wh{P}_{i,j}(1-\wh{P}_{i,j})\emptail}{N_i}}
        + \frac{(5/3)\emptail + \abs{\wh{P}_{i,j}-1/d}}{N_i}
      }
    }^2
    .
    \notag
  \end{equation}
  \label{step:P-bound}

@@

It is easy to combine \cref{thm:err,thm:empirical} to yield intervals
whose widths shrink at least as fast as both the non-empirical
intervals from \cref{thm:err} and the empirical intervals from
\cref{thm:empirical}.
Specifically, determine lower bounds on $\pimin$ and $\gap$ using
\cref{alg:empest},
\[
  \pimin \geq \min_{i \in [d]} \brackets{ \hat\pi_i - \hat{b} }_+
  , \quad
  \gap \geq \brackets{ \hatgap - \hat{w} }_+
  ;
\]
then plug-in these lower bounds for $\pimin$ and $\gap$ in the
deviation bounds in \cref{eq:gapbound} from \cref{thm:err}.
This yields a new interval centered around the estimate of $\gap$ from
\cref{thm:err}, and it no longer depends on unknown quantities.
The interval is a valid $1-2\delta$ probability confidence interval
for $\gap$, and for sufficiently large $n$, the width shrinks at the
rate given in \cref{eq:gapbound}.
We can similarly construct an empirical confidence interval for
$\pimin$ using \cref{eq:piminbound}, which is valid on the same
$1-2\delta$ probability event.\footnote{%
  For the $\pimin$ interval, we only plug-in lower bounds on $\pimin$
  and $\gap$ only where these quantities appear as $1/\pimin$ and
  $1/\gap$ in \cref{eq:piminbound}.
  It is then possible to ``solve'' for observable bounds on $\pimin$.
  See \cref{app:combined} for details.%
}
Finally, we can take the intersection of these new intervals with the
corresponding intervals from \cref{alg:empest}.
This is summarized in the following \lcnamecref{thm:combined}, which
we prove in \cref{app:combined}.
\begin{theorem}
  \label{thm:combined}
  The following holds under the same conditions as
  \cref{thm:empirical}.
  For any $\delta \in (0,1)$, the confidence intervals $\wh{U}$ and
  $\wh{V}$ described above for $\pimin$ and $\gap$, respectively,
  satisfy $\pimin \in \wh{U}$ and $\gap \in \wh{V}$ with probability
  at least $1-2\delta$.
  Furthermore, the widths of these intervals almost surely satisfy
  (as $n \to \infty$) 
  \[
    |\wh{U}|
    =
    O\Parens{
      \sqrt{\frac{\pimin\log\frac{d}{\pimin\delta}}{\gap n}}
    }
    ,
    \quad
    |\wh{V}|
    =
    O\Parens{
      \min\Braces{
        \sqrt{\frac{\log\frac{d}{\delta}\cdot\log(n)}{\pimin\gap n}}
        ,\,
        \hat{w}
      }
    }
  \]
  where $\hat{w}$ is the width from \cref{alg:empest}.
\end{theorem}
